Last glacial cycle glacier erosion potential in the Alps

### 2021-05-26 articles/erosion/response-referee2.md

While we would rather not include such details in the paper, we tried our best
to summarise the effect with this additional sentence in the figure caption:

    Some of the observed basal drag fluctuations in the upper part of the
    transect result from steep slopes on either side of the narrow valley.

>   The three videos are well worth watching  -in fact they clarify some
>   queries arising from the paper. Perhaps they could be captioned in the
>   ‘Video supplement’ paragraph. Some extra information on the third (bedrock
>   altitude) video could point out that there are numerous zero values not
>   shown: the zero geometric mean rates from 2300 to 3000 m around 24 ka
>   otherwise seem to conflict with the numerous positive rates plotted. Zero
>   rates presumably show where ice is frozen to the bed: their altitudinal
>   distribution might be worth discussion in the text. Above 3000 m means are
>   positive, which seems strange.

Well, it seems that you have spend a lot of time looking at our figures and
videos so let us thank you again for that. The caption of Fig. 5 mentioned "the
geometric mean of (non-zero) modelled erosion rates". We found that the
"non-zero" has no reason to be and it was removed. Sorry about the confusion
and please read on for the long explanation.

First, there are no zero values in the animation. As in the paper's Fig. 5, the
animation shows geometric means (i.e. arithmetic means in log-domain), and
geometric means can't contain zero values or they would result in a
divide-by-zero error (or a log-of-zero error). Both animation panels have a
log-scaled x-axis, hence they cannot contain zero values.

On the other hand, both Fig. 5 and the animation depict very low values for
erosion rates (especially around 24 ka). This is due to the glacier physics
embedded in PISM. In particular, the pseudo-plastic sliding law implies that
the glacier is sliding everywhere, but that sliding is infinitely small where
the basal drag is much smaller than the yield stress, including for instance
frozen-bed areas. These tiny sliding velocities yield even tinier erosion rates
(due to the power 2.34 in the erosion law). Such values are not really relevant
and they are off-the-colour-scheme on the Fig. 5 and off-the-chart in the
animation (respectively lower than 10^-9 and 10^-10 mm a^-1).

The only regions where PISM produces truly zero sliding are the regions of zero
surface slopes or zero ice thickness, where the SSA can't be solved. In
practice, this only occurs in the latter case: zero ice thickness. This is the
kind of values we have previously filtered out. Your comment led us to re-check
our calculations. We found that there is no zero-sliding values in the
ice-covered area, and that an ice mask is sufficient to filter out undesirable
values and compute the geometric means. Hence "non-zero" was removed from the
caption.

To be sure, the non-zero filter has been removed from the script generating
companion data (which is used in Fig. 5). This change can be seen here:

TODO: link Github commit.

### 2021-05-26 articles/erosion/alps-erosion.tex
Discuss Koppes et al. (2015) law without outliers.
Remove zero-sliding filter in postproc, caption.

    \begin{figure*}
      \centerline{\includegraphics{alpero_hypsogram}}
      \caption{%
        \textbf{(a)} Potential erosion rate ``hypsogram'', showing the
          geometric mean of (non-zero) modelled rates in 10-m elevation bands
          across the entire model domain and its time evolution. Hatches
          indicate elevation bands with fewer than a hundred grid cells
          (100\,$km^2$).
        \textbf{(b)} Distribution of model domain bedrock topography (grey
          bars) and glaciated topography (blue bars) in 100-m elevation bands,
          and the corresponding cumulative potential erosion volume (brown
          line).
        \textbf{(c)} Same as Fig.~\ref{fig:cumulative}b.
        \todo{fix labels, test hatching below a minimum number of grid cells}}
      \label{fig:hypsogram}
    \end{figure*}

>>>>

    \begin{figure*}
      \centerline{\includegraphics{alpero_hypsogram}}
      \caption{%
        \textbf{(a)} Potential erosion rate ``hypsogram'', showing the
          geometric mean of modelled rates in 10-m elevation bands
          across the entire model domain and its time evolution. Hatches
          indicate elevation bands with fewer than a hundred grid cells
          (100\,$km^2$).
        \textbf{(b)} Distribution of model domain bedrock topography (grey
          bars) and glaciated topography (blue bars) in 100-m elevation bands,
          and the corresponding cumulative potential erosion volume (brown
          line).
        \textbf{(c)} Same as Fig.~\ref{fig:cumulative}b.
        \todo{fix labels, test hatching below a minimum number of grid cells}}
      \label{fig:hypsogram}
    \end{figure*}
